Equity conceptualizations and standards within transportation literature: a scoping review

### 2023-10-05 Manuscript/Manuscript.qmd
Antonio is still editing

The fiery wake of a rocket could be seen ascending, piercing the blue skies of West Texas. It was mid-morning on July 20, 2021, and on board was a small group of four passengers that included Jeff Bezos, the founder of Amazon and then the world's richest person [@harwoodBezos2021]. The mission that morning was among the first ever crewed private suborbital flights, and the adventure (described as "intense" by a member of the crew) lasted a total of 10 minutes and 10 seconds [@harwoodBezos2021]. In addition to intense, the undertaking was expensive: a seat for the flight had previously been auctioned for no less than $28 million USD [@griffinPerson2021]. Before even 10:00 am (EDT) Bezos was back from his excursion, and right on time to declare that this was the "[b]est day ever". To reporters covering the event he said "I want to thank every Amazon employee and every Amazon customer, because you guys paid for all of this" [@johnsonRoadToSpace2021].

@@

Multiple national and cross-national studies suggest that people in many societies have some tolerance for inequality [@kiatpongsanHow2014]: indeed, it appears that _some_ stratification, as suggested by @davisSome1945, is thought to fulfill a valuable function. However, extreme inequality is often frowned upon, and can lead to social dysfunction and other ills [@acemogluWhy2000; @taydasCan2012; @duIncome2019; @houleInequality2022]. But the perceptions of what is "fair" are neither universal or static. Instead, they are malleable, and can be affected by the existence of opportunities for social mobility [@shariffIncome2016; @artigeWhat2019], by exposure to inequality [@schroderIncome2017; @garciaChanging2023], by learned helplessness [@kimSocioeconomic2022], and even by how information about inequality is communicated to the public [@walkerPeople2021]. It follows that, in general terms, there are at least three different manners of thinking about inequality: 1) in a _positive_ (or descriptive) manner, as the current state of the distribution of benefits and burdens of things; 2) also descriptively, as the _perceptions_ about the distribution of those benefits and burdens; and 3) normatively (or prescriptively), as the desired or ideal state of the same. Clearly, the three may coincide (e.g, if the perceived levels of inequality matched actual inequality and also how much of it the public desired). However, they do not necessarily have to, and in many cases will differ from one another. Measuring inequality as it is, in the most objective way possible is an essential task to decide whether there is a need to develop inequality-related policy to increase fairness. In turn, measuring the perceptions of inequality, in the most accurate way possible, may be important to achieve sufficient public buy-in in order to enhance the chances that a given policy will succeed.

@@

Supporting the creation of (more) equitable transportation systems involves understanding the production, distribution, and management of transportation benefits and costs; how they are distributed; and what values are implemented (and by whom) in the form of standards [@pereiraDistributiveJusticeEquity2017a; @shellerMobilityJusticePolitics2018; @pereiraTransportationEquity2021]. Thus, for this review we engage the literature with the following questions in mind: 1) what is our current understanding of the things that transportation systems do, for whom, and to whom; 2) what does the literature say about the distribution of the benefits and costs of transportation systems; 3) and what values are embodied in normative statements about said distribution. Ultimately, this review aims to collate the academic knowledge and present it in a manner useful to support both academics and decision-makers in the development and implementation of standards for equity in transportation planning and policy, some of which are sorely missing, even in those places where legislation explicitly calls for justice^[[As recently as -@martensFair2021, Martens and Golub note that Federal directives related to Title VI in the US "do not provide guidelines that can help agencies develop explicit standards to assess the distribution of accessibility benefits from projects or plans."]].

@@

Several forms of justice can be distinguished, including [see @jaggarPhilosophicalChallengesGlobal2009; @karnerTransportationEquityTransportation2020]:

@@

**Reparative justice**. This form of justice relates to the proper way to correct or rectify past wrongs even if the wrongdoers can no longer be meted retribution.

@@

Fairness, in contrast to equity, is more complicated to define. The concept does not have the same history of development as an analytical tool, and can be interpreted in numerous, and possibly discordant ways. That this is the case is convincingly demonstrated by @martensFair2021 in their study of the application of Title VI of the Civil Rights Act of 1964 in accessibility planning practice in the US. Title VI explicitly talks about the distribution of benefits derived from Federal funding: "[n]o person in the United States shall, on the ground of race, color, or national origin, be excluded from participation in, be denied the benefits of, or be subjected to discrimination under any program or activity receiving Federal financial assistance." However, as @martensFair2021 show, there are several ways to comply with regulations while achieving different outcomes: from a very mild form of fairness that simply avoids intentional discrimination, to strategically overdeliver the benefits to compensate for past discrimination (reparative justice). What justice did fairness serve in each case? It depends on what was the reason for seeking justice in the first place! Our reading of @martensFair2021 is that fairness is a yardstick that is best used _a priori_ than _a posteriori_, for doing so risks rationalizing the outcomes rather than

The focus of this report is on standards of fairness that, combined with the use of equity analysis, can help understand how better to move towards just outcomes in transportation planning.

@@

For the analysis we are inspired by the framing of @jaggarPhilosophicalChallengesGlobal2009 for philosophical questions of justice.

borrow a "Where?", "When?", "Who?", "What?", and "How?" framing to categorize aspects of how equity is assessed.

<!-- The texture of these questions becomes more crisp when talking about transportation, which is inherently about space and time, that is to say about "where" and "when". Conventionally, theories of justice have been aspatial, despite an early interest of geographers on matters of justice [@]-->

This categorization is outlined in  to describe substantive sets of answers to questions of distributive justice (and further discussed in the context of justice within the critical human geography discipline [@przybylinskiWhereJusticeGeography2022]).
For questions of *justice*, convincing answers to these questions require a rationale: a "Why?" [@jaggarPhilosophicalChallengesGlobal2009].
For the purpose of this review focused on *equity*, the focus is on studies that substantively answer "Where?", "When?", "Who?", "What?", and "How?"; the answer to "Why?" may be incidental.

@@

To understand the distributional implications of transportation systems, it is essential that we are clear about what they do. In other words, what do transportation systems _produce_?

At their most fundamental, transportation systems are space-time convergence tools, tools that improve the rate at which time is traded for space. They usually do this by increasing the speed of movement: sidewalks facilitate walking, traffic lights facilitate the ordered flow of vehicular traffic, and a launching pad makes it possible for a rocket to takeoff. With complex interlocking parts (sidewalk, road, traffic sign, parking regulations), transportation systems produce _mobility_, the potential for movement. Combined with land use systems (the spatial distribution of opportunities on the landscape), they produce _accessibility_, the potential to reach destinations. The realization of those potentials happens through travel. But, as the adage goes, travel is derived demand, which seems to hold for most (even if not all) travel.

>>>>

It was a common kind of morning in West Texas but for the fiery wake of a rocket that could be seen ascending, piercing the blue skies. The date was July 20, 2021, and on board was a small group of four passengers that included Jeff Bezos, the founder of Amazon and then the world's richest person [@harwoodBezos2021]. The mission that morning was among the first ever crewed private suborbital flights, and the adventure (described as "intense" by a member of the crew) lasted a total of 10 minutes and 10 seconds [@harwoodBezos2021]. In addition to intense, the undertaking was expensive: a seat for the flight had previously been auctioned for no less than $28 million USD [@griffinPerson2021]. Before even 10:00 am (EDT) Bezos was back from his excursion, and right on time to declare that this was the "[b]est day ever". To reporters covering the event he said "I want to thank every Amazon employee and every Amazon customer, because you guys paid for all of this" [@johnsonRoadToSpace2021].

@@

Multiple national and cross-national studies suggest that people in many societies have some tolerance for inequality [@kiatpongsanHow2014]: indeed, it appears that _some_ stratification, as suggested by @davisSome1945, is thought to fulfill a valuable function. However, extreme inequality is often frowned upon, and can lead to social dysfunction and other ills [@acemogluWhy2000; @taydasCan2012; @duIncome2019; @houleInequality2022]. But the perceptions of what is "fair" are neither universal or static. Instead, they are malleable, and can be affected by the existence of opportunities for social mobility [@shariffIncome2016; @artigeWhat2019], by exposure to inequality [@schroderIncome2017; @garciaChanging2023], by learned helplessness [@kimSocioeconomic2022], and even by how information about inequality is communicated to the public [@walkerPeople2021]. It follows that, in general terms, there are at least three different manners of thinking about inequality^[@martensTransport2016, in the introduction of his landmark text, talks about explanatory and prescriptive theories of justice; despite the saying "the arc of history is long and bends toward justice", it is unlikely that a predictive theory of justice exists.]: 1) in a _positive_ (or descriptive) manner, as the current state of the distribution of benefits and burdens of things; 2) also descriptively, as the _perceptions_ about the distribution of those benefits and burdens; and 3) normatively (or prescriptively), as the desired or ideal state of the same. Clearly, the three may coincide (e.g, if the perceived levels of inequality matched actual inequality and also how much of it the public desired). However, they do not necessarily have to, and in many cases will differ from one another. Measuring inequality as it is, in the most objective way possible is an essential task to decide whether there is a need to develop inequality-related policy to increase fairness. In turn, measuring the perceptions of inequality, in the most accurate way possible, may be important to achieve sufficient public buy-in in order to enhance the chances that a given policy will succeed.

@@

Supporting the creation of (more) just transportation systems involves understanding the production, distribution, and management of transportation benefits and costs; how they are distributed; and what values are implemented (and by whom) in the form of standards [@pereiraDistributiveJusticeEquity2017a; @shellerMobilityJusticePolitics2018; @pereiraTransportationEquity2021]. Thus, for this review we engage the literature with the following questions in mind: 1) what is our current understanding of the things that transportation systems do, for whom, and to whom; 2) what does the literature say about the distribution of the benefits and costs of transportation systems; 3) and what values are embodied in normative statements about said distribution. Ultimately, this review aims to collate the academic knowledge and present it in a manner useful to support both academics and decision-makers in the development and implementation of standards for equity in transportation planning and policy. In this, we aim to provide relief even, or particularly, in those places where legislation explicitly calls for justice^[[As recently as -@martensFair2021, Martens and Golub note that Federal directives related to Title VI in the US "do not provide guidelines that can help agencies develop explicit standards to assess the distribution of accessibility benefits from projects or plans."]].

@@

Several forms of justice can be distinguished, including [see @jaggarPhilosophicalChallengesGlobal2009; @pereiraDistributiveJusticeEquity2017a; @karnerTransportationEquityTransportation2020]:

@@

**Reparative (or restorative) justice**. This form of justice relates to the proper way to correct or rectify past wrongs even if the wrongdoers can no longer be meted retribution.

@@

Fairness, in contrast to equity, is more complicated to define. The concept does not have the same history of development as an analytical tool, and can be interpreted in numerous, and possibly discordant ways. That this is the case is convincingly demonstrated by @martensFair2021 in their study of the application of Title VI of the Civil Rights Act of 1964 in accessibility planning practice in the US. Title VI explicitly talks about the distribution of benefits derived from Federal funding: "[n]o person in the United States shall, on the ground of race, color, or national origin, be excluded from participation in, be denied the benefits of, or be subjected to discrimination under any program or activity receiving Federal financial assistance." However, as @martensFair2021 show, there are several ways to comply with regulations while achieving different outcomes: from the banal (do not _knowingly_ discriminate), to the substantive (compensation for past discrimination, i.e., reparative justice). What justice does fairness serve in each case? It depends on what was the reason for seeking justice in the first place! Our reading of @martensFair2021 is that fairness is a yardstick that is best used _a priori_ than _a posteriori_, for doing so risks rationalizing the outcomes rather than

The focus of this report is on standards of fairness that, combined with the use of equity analysis, can help understand how better to move towards just transportation outcomes.

Here we are inspired by the framing of @jaggarPhilosophicalChallengesGlobal2009 for philosophical questions of justice^[Similar questions are found peppered throughout the literature either explicity, as for example in @karnerTransportationEquityTransportation2020, who ask "of what", "for whom," and "how much" in reference to equity, or implicitly, as in @gosslingUrban2016, who asks "what?" (exposure, space, access) and "for whom?" (gender, age, ethnicity) about the outputs of transportation.]. According to -@jaggarPhilosophicalChallengesGlobal2009, Western philosophy has approached the issue of justice by asking "Where?", "When?", "Who?", "What?", and "How?". Conventionally, discussions about justice have been aspatial, or rather, seen from the point of view of social space instead of geographical space, despite an early interest of geographers on the matter [@pirieSpatial1983]. The texture of the questions becomes more immediate and crisp when talking about transportation, which is inherently about space and time.

- **"Where?"** Traditionally related to the applicable domain or sphere of life relevant for justice. Conventionally this meant the in-group [e.g., members of the same nation state; @jaggarPhilosophicalChallengesGlobal2009, p. 3], a notion since expanded (e.g., with the Universal Declaration of Human Rights) to include all members of the human species, but has yet to cover other living creatures.

@@

- "When?" asks what are the social circumstances within which the
demands of justice have application.
- "Who?" asks which entities should be regarded as subjects of justice,
meaning who or what are entitled to make justice claims deserving of
moral consideration.
- "What?" asks which entities should be regarded as objects of justice,
meaning which kinds or categories of things should be distributed in a
just manner.
- Finally, "How?" asks which principles are the most morally appropriate
for guiding the allocation of various objects to various subjects in various
circumstances.

This categorization is outlined in  to describe substantive sets of answers to questions of distributive justice (and further discussed in the context of justice within the critical human geography discipline [@przybylinskiWhereJusticeGeography2022]).

For questions of *justice*, convincing answers to these questions require a rationale: a "Why?" [@jaggarPhilosophicalChallengesGlobal2009].
For the purpose of this review focused on *equity*, the focus is on studies that substantively answer "Where?", "When?", "Who?", "What?", and "How?"; the answer to "Why?" may be incidental.

@@

To understand the distributional implications of transportation systems, it is essential that we are clear about what they do. What do transportation systems _produce_?

At their most fundamental, transportation systems are space-time convergence tools, tools that improve the rate at which time is traded for space. They usually do this by increasing the speed of movement: sidewalks facilitate walking, traffic lights facilitate the ordered flow of vehicular traffic, and a launching pad makes it possible for a rocket to takeoff. With complex interlocking parts (sidewalk, road, traffic sign, parking regulations), transportation systems produce _mobility_, the potential for movement. Combined with land use systems (the spatial distribution of opportunities on the landscape), they produce _accessibility_, the potential to reach destinations. The realization of those potentials happens through travel. But, as the adage goes, travel is derived demand, which seems to hold for most (even if not all) situations [ref mokhtarian].
